---
title: "GP40 LSA"
author: "Yamé Meyvaert"
date: "2024-11-28"
output: html_document
---

## Loading the data

Installing and loading the libraries:

```{r}
#install.packages("tidyverse")
library(tidyverse)
library(readxl)
```

```{r}
cancer_dataset <- read_csv('metadata.csv')
cancer_abundances <- read_excel('abundances.xlsx')
```

###Data exploration Metadata

Checking datatypes:
```{r}
str(cancer_dataset)
```

Converting datatypes (cleaning the data):
```{r}
cancer_dataset$Gender <- as.factor(cancer_dataset$Gender)
cancer_dataset$Proteomic_Subtype <- as.factor(cancer_dataset$Proteomic_Subtype)
cancer_dataset$`Diameter_of_tumor_(cm)` <- as.numeric(cancer_dataset$`Diameter_of_tumor_(cm)`)
cancer_dataset$Survival_status <- as.factor(cancer_dataset$Survival_status)
cancer_dataset$Recurr_status <- as.factor(cancer_dataset$Recurr_status)

#checking for correct converting
str(cancer_dataset)
```

Check the header
```{r}
head(cancer_dataset)
```

Checking for missing values:
```{r}
missing_values <- colSums(is.na(cancer_dataset))
print(missing_values)
```

Interpretation: 7 missing values in Diameter_of_tumor_(cm) column and 10 in Tumor_cellularity_(%).

Missing data visualisation:
```{r}
#install.packages("naniar")
library(naniar)

vis_miss(cancer_dataset)       # Heatmap van missende waarden
gg_miss_var(cancer_dataset)    # Barplot van missende waarden per kolom
```

Basic stats for each of the columns:
```{r}
summary(cancer_dataset)
```
```{r}
#install.packages("skimr")
library(skimr)
skim(cancer_dataset)
```

Distribution of the data
```{r}
cancer_dataset %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, min = min, max = max), na.rm = TRUE))
```

```{r}
cancer_dataset%>%
  count(Proteomic_Subtype)
```

###Data exploration Abundances data

```{r}
str(cancer_abundances)
```

Check the header
```{r}
head(cancer_abundances)
```

```{r}
cancer_abundances %>%
  mutate(missing_count = rowSums(is.na(.)))%>%
  select(ID, missing_count)
```

```{r}
summary(cancer_abundances)
```


Hypothese 1:

Hypotheses:
Null hypothesis (H₀): The median Tumor Cellularity is the same across all proteomic subtypes. 
Alternative hypothesis (Hₐ): At least one Proteomic Subtype has a different median Tumor Cellularity. 

Missing values:
```{r}
vis_miss(cancer_dataset)       # Heatmap of missing values
gg_miss_var(cancer_dataset)    # Barplot of missing values
```
Check how many NA values in Tumor cellularity:
```{r}
sum(is.na(cancer_dataset$`Tumor_cellularity_(%)`))
```
Random? Or not random?
```{r}
table(is.na(cancer_dataset$Proteomic_Subtype), is.na(cancer_dataset$`Tumor_cellularity_(%)`))

# Contingency table for missingness
missing_table <- table(is.na(cancer_dataset$Proteomic_Subtype), is.na(cancer_dataset$`Tumor_cellularity_(%)`))

# Chi-squared test
chisq.test(missing_table)

```
Interpretation: the p-value is below 0.05 which means that the missingness of the values is not random and therefore we can't remove them from the column. 

```{r}
#install.packages("naniar")
library(naniar)

#Plot Missingness:

# Check missing data patterns
vis_miss(cancer_dataset)  

```

Replace NA with median:
```{r}
# Calculate the median while ignoring NA values
median_value <- median(cancer_dataset$`Tumor_cellularity_(%)`, na.rm = TRUE)

# Replace NA values with the calculated median
cancer_dataset$`Tumor_cellularity_(%)`[is.na(cancer_dataset$`Tumor_cellularity_(%)`)] <- median_value

# Print the column to check if NAs were replaced
print(cancer_dataset$`Tumor_cellularity_(%)`)

```

Standardization:
```{r}
cancer_dataset$standardized_cellularity <- scale(cancer_dataset$`Tumor_cellularity_(%)`, center = TRUE, scale = TRUE)

# Visualize the standardized column
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = standardized_cellularity)) +
  geom_boxplot(fill = "grey") +
  labs(
    title = "Boxplot of Standardized Tumor Cellularity (%) by Proteomic Subtype",
    x = "Proteomic Subtype",
    y = "Standardized Tumor Cellularity (z-score)"
  ) +
  theme_minimal()

#visualize without standardization:
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Tumor Cellularity by Proteomic Subtype", y = "Tumor Cellularity (%)", x = "Proteomic Subtype")

```
Standardization is not needed because Tumor cellularity is in the same scale in the 3 different groups of the proteomic subtype.

Filter out outliers:
```{r}
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot(outlier.color = "purple", outlier.shape = 8) +
  theme_minimal() +
  labs(title = "Tumor Cellularity by Proteomic Subtype", y = "Tumor Cellularity (%)", x = "Proteomic Subtype")

```

```{r}
# Calculate IQR-based thresholds
Q1 <- quantile(cancer_dataset$`Tumor_cellularity_(%)`, 0.25)
Q3 <- quantile(cancer_dataset$`Tumor_cellularity_(%)`, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter data based on IQR thresholds
filtered_dataset <- cancer_dataset %>%
  filter(`Tumor_cellularity_(%)` >= lower_bound & `Tumor_cellularity_(%)` <= upper_bound)

# Create boxplot with the filtered data
ggplot(filtered_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot(outlier.color = "purple", outlier.shape = 8) +
  labs(
    title = "Boxplot of Tumor Cellularity (%) by Proteomic Subtype (IQR Method)",
    x = "Proteomic Subtype",
    y = "Tumor Cellularity (%)"
  ) +
  theme_minimal()

```
There were outliers in the dataset. These will be filtered out. 


###Data visualisation

Histogram for numerical values
```{r}
ggplot(filtered_dataset, aes(x = `Tumor_cellularity_(%)`)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  theme_minimal()
```

```{r}
filtered_dataset %>% ggplot(aes(x = `Tumor_cellularity_(%)` , fill = Proteomic_Subtype)) + geom_histogram() +theme_classic()
```

Boxplot:
```{r}
ggplot(filtered_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal()
```
Normal distribution per Subtype:
```{r}
library(dplyr)

# Perform Shapiro-Wilk test for normality on each subtype
normality_results <- filtered_dataset %>%
  filter(!is.na(`Tumor_cellularity_(%)`)) %>%  # Remove rows with NA values
  group_by(Proteomic_Subtype) %>%             # Group by subtype
  summarise(
    shapiro_p_value = shapiro.test(`Tumor_cellularity_(%)`)$p.value
  )

print(normality_results)

```
A small p-value (< 0.05) suggests the data is not normally distributed.


Visualisation:
```{r}
ggplot(filtered_dataset, aes(x = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Tumor Cellularity by Subtype", 
       x = "Tumor Cellularity (%)",
       y = "Density") +
  theme_minimal()
```

Tumor cellullarity data are not normally ditributed per Proteomic Subtype

```{r}


ggplot(filtered_dataset, aes(x = log(`Tumor_cellularity_(%)` + 1), fill = Proteomic_Subtype)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density Plot of Log-Transformed Tumor Cellularity by Subtype", 
    x = "Log(Tumor Cellularity (%) + 1)",
    y = "Density"
  ) +
  theme_minimal()
```
Since the log transformation did not result in approximate normality, nonparametric statistical methods that do not assume normality will be considered for further analysis.

P-values testing: 

Kruskal-Wallis test
```{r}
kruskal.test(`Tumor_cellularity_(%)` ~ Proteomic_Subtype, data = filtered_dataset)
```
Interpretation: With a p-value of 
0.01095, which is less than the common significance level (α=0.05), the test result is statistically significant.
This means that there is strong evidence to reject the null hypothesis.
At least one of the Proteomic_Subtype groups has a significantly different median Tumor_cellularity_(%).

Since the Kruskal-Wallis test is an omnibus test, it doesn't specify which groups differ. To identify the specific groups with differing medians, you can perform pairwise post-hoc tests: (post-hoc tests are needed to specify which groups differ)

```{r}
pairwise.wilcox.test(filtered_dataset$`Tumor_cellularity_(%)`, 
                     filtered_dataset$Proteomic_Subtype, 
                     p.adjust.method = "bonferroni")


#Visualisation:
ggplot(filtered_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Tumor Cellularity by Proteomic Subtype", y = "Tumor Cellularity (%)")

```
Interpretation: Results Interpretation
- S-I vs S-II:
Adjusted p-value = 0.763.
There is no statistically significant difference in the median Tumor_cellularity_(%) between subtypes S-I and S-II.

- S-I vs S-III:
Adjusted p-value = 0.142.
This is not significant at the 0.05 level. You might consider the difference to be marginal or weak evidence against the null hypothesis.

- S-II vs S-III:
Adjusted p-value = 0.011.
This indicates a statistically significant difference in the median Tumor_cellularity_(%) between subtypes S-II and S-III at the 0.05 level.

The warning occurs due to ties in Tumor_cellularity_(%), which is common with real-world data. This doesn't invalidate your results:

Monte Carlo approximation for more robust p-values:
```{r}
pairwise.wilcox.test(filtered_dataset$`Tumor_cellularity_(%)`, 
                     filtered_dataset$Proteomic_Subtype, 
                     p.adjust.method = "bonferroni", 
                     exact = FALSE)

```
With the Monte-Carlo approximation there is no difference in p-values. When ties are few and the data size is reasonable, the normal approximation (used in exact = FALSE) provides a good estimate of the p-values. Therefore thee approximate p-values provided by the pairwise Wilcox test are reliable and consistent with the results generated via Monte Carlo simulations.

The Wilcoxon test is less sensitive to outliers.


Hypothese 2:

H₀: Age, gender, and Disease-free survival have no statistically significant effect on Recurrence Status.  
Hₐ: At least one of the independent variables (age, gender, or Disease-free survival) has a statistically significant effect on Recurrence Status. 

Check missing values:
```{r}
#install.packages("naniar")
library(naniar)

#Plot Missingness:

# Check missing data patterns
vis_miss(cancer_dataset)  # naniar package

```
There are no missing values in Age, gender, disease free survival and recurrence status

Filter out oultiers:
```{r}
#without outlier check
ggplot(cancer_dataset, aes(x = Recurr_status, y = age)) +
  geom_boxplot(fill = "lightpink") +
  theme_minimal()
#with outlier check:
# Calculate IQR-based thresholds for age
Q1 <- quantile(cancer_dataset$age, 0.25)
Q3 <- quantile(cancer_dataset$age, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter data based on IQR thresholds
filtered_dataset_age <- cancer_dataset %>%
  filter(age >= lower_bound & age <= upper_bound)

# Create boxplot with the filtered data
ggplot(filtered_dataset_age, aes(x = Recurr_status, y = age)) +
  geom_boxplot(fill = "darkgreen") +
  labs(
    title = "Boxplot of Age by Recurrence Status (IQR Method)",
    x = "Recurrence Status",
    y = "Age"
  ) +
  theme_minimal()

```

```{r}
#without outlier check:
ggplot(cancer_dataset, aes(x = Recurr_status, y = `Disease_free_survival_(m)`)) +
  geom_boxplot(fill = "lightpink") +
  theme_minimal()

#with outlier check 
# Calculate IQR-based thresholds for Disease_free_survival_(m)
Q1 <- quantile(cancer_dataset$`Disease_free_survival_(m)`, 0.25)
Q3 <- quantile(cancer_dataset$`Disease_free_survival_(m)`, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter data based on IQR thresholds
filtered_dataset3 <- cancer_dataset %>%
  filter(`Disease_free_survival_(m)` >= lower_bound & `Disease_free_survival_(m)` <= upper_bound)

# Create boxplot with the filtered data
ggplot(filtered_dataset3, aes(x = Recurr_status, y = `Disease_free_survival_(m)`)) +
  geom_boxplot(fill = "darkblue") +
  labs(
    title = "Boxplot of Disease-Free Survival (months) by Recurrence Status (IQR Method)",
    x = "Recurrence Status",
    y = "Disease-Free Survival (months)"
  ) +
  theme_minimal()

```

```{r}
ggplot(cancer_dataset, aes(x = Recurr_status, y = age, fill = Recurr_status)) +
  geom_boxplot(outlier.color = "purple", outlier.shape = 8) +
  theme_minimal() +
  labs(title = "Age by Recurrence status", y = "Age", x = "Recurrence status")
```

```{r}
ggplot(cancer_dataset, aes(x = Recurr_status, y = `Disease_free_survival_(m)`, fill = Recurr_status)) +
  geom_boxplot(outlier.color = "purple", outlier.shape = 8) +
  theme_minimal() +
  labs(title = "Disease free survival by Recurrence status", y = "Age", x = "Recurrence status")
```

There are outliers present in the variable age. These will be filtered out. But no outliers were detected for Disease_free_survival_(m).

Distribution:
```{r}
ggplot(filtered_dataset_age, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  theme_minimal()
```

```{r}
library(ggpubr)
ggqqplot(filtered_dataset_age$age)+
  ggtitle("Age")
```
P-values testing with Shapiro for normal distribution:
```{r}
filtered_dataset_age$age <- as.numeric(as.character(filtered_dataset_age$age))

shapiro.test(filtered_dataset_age$age)
```
The results of the Shapiro-Wilk test indicate the following for the variable age:

Test Statistic (W): 
W=0.99226
The closer this value is to 1, the more the data resembles a normal distribution.

p-value: 
p=0.844

A p-value > 0.05 suggests that we fail to reject the null hypothesis, which means there is no significant evidence to suggest that the data is not normally distributed.

Interpretation:
Based on these results, the age data in your cancer_dataset appears to follow a normal distribution. While this doesn't guarantee perfect normality, it suggests that deviations from normality are not statistically significant.

```{r}
library(ggpubr)
ggqqplot(filtered_dataset_age$`Disease_free_survival_(m)`)+
  ggtitle("Disease free survival")

```
The data for Disease_free_survival_(m) is not normally distributed.

P-values testing with Shapiro for normal distribution:
```{r}
filtered_dataset_age$`Disease_free_survival_(m)` <- as.numeric(as.character(filtered_dataset_age$`Disease_free_survival_(m)`))

shapiro.test(filtered_dataset_age$`Disease_free_survival_(m)`)
```
The results of Shapiro-Wilk test indicate the following for the variable Disease_free_survival_(m):

Test Statistic (W):
W=0.81088
The farther this value is from 1, the less the data resembles a normal distribution.

p-value:
p=6.216×10^−10
 
A p-value < 0.05 suggests that we reject the null hypothesis, meaning there is significant evidence to indicate that the data does not follow a normal distribution.

Interpretation:
Based on these results, the Disease_free_survival_(m) data in the filtered_dataset_age does not follow a normal distribution. The significant deviation from normality implies that statistical methods relying on the assumption of normality may not be appropriate for this variable.

Histograms:
```{r}
ggplot(data = filtered_dataset_age, aes(x = age,
                     fill = Recurr_status))+
  geom_histogram()+
  theme_classic()
```
```{r}
ggplot(data = filtered_dataset_age, aes(x = `Disease_free_survival_(m)`,
                     fill = Recurr_status))+
  geom_histogram()+
  theme_classic()
```
The data does not need to be normally distributed for the logistic regression model. 

Logistic model:
```{r}
blm_recurr <- glm(Recurr_status ~ age+Gender+`Disease_free_survival_(m)`, data=filtered_dataset_age, family="binomial" )

summary(blm_recurr)
```
Based on the p-values only the parameter Disease free survival is significant, as it has a value below the 0.05 treshold.

Odds ratio:
```{r}
odds <- exp(coef(blm_recurr))
odds
```
Interpretation:
Odds Ratio for Disease_free_survival_(m):
OR=0.9643 (as per your output).This indicates that for every one-unit increase in Disease-free survival (in months), the odds of recurrence decrease by about 1−0.964 ≈3.6%, holding all other variables constant.

Since OR<1, longer disease-free survival is associated with reduced odds of recurrence.

```{r}
#install.packages("forestmodel")
library(forestmodel)
forest_model(blm_recurr)
```
With these results there is a possibility to make a Kaplan-Meier plot with disease free survival and the recurrence status:
```{r}
#install.packages(c("survival","survminer"))
library(survival)
library(survminer)
```

```{r}
cancer_dataset %>%
  group_by(Recurr_status) %>%
  summarise(n())
```

Kaplan-Meier plot:
```{r}
# create the surv object
surv_object <- Surv(filtered_dataset_age$`Disease_free_survival_(m)`, as.numeric(filtered_dataset_age$Recurr_status))

# fit a KM curve to the data
fit <- survfit(surv_object ~ Recurr_status, data=filtered_dataset_age)
summary(fit)
```
```{r}
# plot this curve
ggsurvplot(fit, data = filtered_dataset_age,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_classic()) # Change ggplot2 theme
```
Log rank test:
```{r}
fitdif <- survdiff(surv_object ~ Recurr_status, data=filtered_dataset_age)
fitdif
```

Interpretation:
In first instance we could look to the p-value that comes out of the survdif analysis. In this case the p-value is lower than 2e-16. We could thus accept or alternative hypothesis and state that the two groups have a different Disease free survival. We could also look to the plot we made with ggsurvplot(). The confidence intervals are not overlapping, and the p-value is lower than 0.001. Both are an indication that the two groups have indeed a different Disease free survival. At the moment we know that there are differences, but how big is this difference? You might get an indication from your plot, by reading the median survival time. Which is the moment where 50% of the patients in that group died. We can see that this is not reached for the non-recurrence patients and its about 10 months for the Recurrence patients.


We could also look at if square root transformation would improve the logistic regression model: 
-> One assumption of this model is the variables are linear with the log odds of the outcome

```{r}
install.packages("ResourceSelection")
library(ResourceSelection)
# Add predictions from the model
filtered_dataset_age$predicted <- predict(blm_recurrence, type = "response")
# Bin the data
filtered_dataset_age$bin <- cut(filtered_dataset_age$predicted, breaks = 10)
# Plot the mean of predictors against log-odds of outcome
logit <- log(filtered_dataset_age$predicted / (1 - filtered_dataset_age$predicted))
plot(filtered_dataset_age$`Disease_free_survival_(m)`, logit, 
      main = "Disease Free Survival vs Log-Odds") # Example for 'Disease_free_survival_(m)'

# Step 2: Apply square root transformation
filtered_dataset_age$`Disease_free_survival_(m)_sqrt` <- sqrt(filtered_dataset_age$`Disease_free_survival_(m)`)

# Fit the logistic regression model with transformed 'Disease_free_survival_(m)_sqrt'
blm_recurr_sqrt1 <- glm(Recurr_status ~ age + Gender + `Disease_free_survival_(m)_sqrt`, 
                       data = filtered_dataset_age, family = "binomial")

# Get the predicted values from the model
filtered_dataset_age$predicted_sqrt <- predict(blm_recurr_sqrt1, type = "response")

# Compute the log-odds
logit_sqrt <- log(filtered_dataset_age$predicted_sqrt / (1 - filtered_dataset_age$predicted_sqrt))

# Plot the relationship between the transformed 'Disease_free_survival_(m)_sqrt' and log-odds
plot(filtered_dataset_age$`Disease_free_survival_(m)_sqrt`, logit_sqrt, 
     main = "Square Root of Disease Free Survival vs Log-Odds")
```


```{r}
#install.packages("ResourceSelection")
library(ResourceSelection)
# Add predictions from the model
filtered_dataset_age$predicted <- predict(blm_recurrence, type = "response")
# Bin the data
filtered_dataset_age$bin <- cut(filtered_dataset_age$predicted, breaks = 10)
# Plot the mean of predictors against log-odds of outcome
logit <- log(filtered_dataset_age$predicted / (1 - filtered_dataset_age$predicted))
plot(filtered_dataset_age$`age`, logit, 
      main = "age vs Log-Odds") # Example for 'age'


# Step 2: Apply square root transformation
filtered_dataset_age$`age_sqrt` <- sqrt(filtered_dataset_age$`age`)

# Fit the logistic regression model with transformed 'age_sqrt'
blm_recurr_sqrt1 <- glm(Recurr_status ~ age + Gender + `age_sqrt`, 
                       data = filtered_dataset_age, family = "binomial")

# Get the predicted values from the model
filtered_dataset_age$predicted_sqrt <- predict(blm_recurr_sqrt1, type = "response")

# Compute the log-odds
logit_sqrt <- log(filtered_dataset_age$predicted_sqrt / (1 - filtered_dataset_age$predicted_sqrt))

# Plot the relationship between the transformed 'Disease_free_survival_(m)_sqrt' and log-odds
plot(filtered_dataset_age$`age_sqrt`, logit_sqrt, 
     main = "Square Root of age vs Log-Odds")
```

For the variable age, there is no improvement of this transformation on the linearity. 
```{r}
blm_recurrence <- glm(Recurr_status ~ age+Gender+`Disease_free_survival_(m)`, data=filtered_dataset_age, family="binomial" )
summary(blm_recurrence)

blm_recurr_sqrt1 <- glm(Recurr_status ~ age + Gender + `Disease_free_survival_(m)_sqrt`, 
                       data = filtered_dataset_age, family = "binomial")
summary(blm_recurr_sqrt1)
```

=> New model with the square root transformation of Disease free survival is better! You have lower AIC and residual deviance, indicating a better overall fit!
The p-value for the square root-transformed Disease_free_survival_(m) is lower, suggesting a stronger statistical association with Recurrence Status. This indicates that the transformation helped linearize the relationship between Disease-free Survival and Recurrence Status, making the logistic regression assumptions more appropriate.

```{r}
odds <- exp(coef(blm_recurr_sqrt1))
odds
```

```{r}
#install.packages("forestmodel")
library(forestmodel)
forest_model(blm_recurr_sqrt1)
```

Before Transformation (OR = 0.9643):

The odds ratio of 0.9643 indicates that for every one-month increase in Disease_free_survival_(m), the odds of recurrence decrease by approximately 3.6% (1 - 0.9643 = 0.036 or 3.6%).
This effect size suggests that longer disease-free survival is associated with a lower risk of recurrence, though the effect per month is relatively small.
After Square Root Transformation (OR = 0.5660):

The odds ratio of 0.5660 indicates that for every one-unit increase in the square root of Disease_free_survival_(m), the odds of recurrence decrease by approximately 43.4% (1 - 0.5660 = 0.434 or 43.4%).
Since the scale has changed, the interpretation is now based on the square root of the survival time, meaning the effect of disease-free survival on recurrence is stronger and non-linear.


